---
title: "Clean 1 - Medata curation"
author: "Tal Gavriel"
date: "2025-03-09"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
# In this script:

final curation of Medata and trait data

*Input data:*


* `Clean_Medata_lis.Rdata` - the Medata after DD script cleaning

* `env_var.Rdata` -  updated bio-orcal environmental data 

* `trait_data_cleaned.csv`- traits data from Avary

* `trait_imputed.csv`- imputed traits data from Belmaker lab FD hackathon 

* `ab_full.csv` -  conversion coefficient table

* `missing_commercial_full.csv`- species commercial value

*  all data stored in the `1_raw_input_data` folder

*Output data:*

* `medata_multistressor.rdata` - medata clean for this project

* `species_traits_multistressor.rdata` - trait data clean for this project

* `site_biomass.rdata` - table of the native and total biomass per site

* `site_biomass_commercial.rdata` - native and invasive biomass per site & commercial value 

* `final_commercial_list.rdata`- the final list of the species commercial values

*  all output saved in the `2_data_for_analyses` folder



### Load packeges

```{r message=FALSE, warning=FALSE}
library(iNEXT.3D)
library(tidyverse)
library(vegan)
library(plotrix)
library(sjPlot)
library(sjmisc)
library(gridExtra)
library(ggeffects)
library(vegan)
library(plotly)
library(FD)
library(mFD)
library(funrar)
library(here)

select<-dplyr::select

```



# load data

```{r message=FALSE, warning=FALSE}

load(here("1_raw_input_data", "Clean_Medata_lis.Rdata"))
load(here("1_raw_input_data", "env_var.Rdata"))

ab_full <- read_csv(here("1_raw_input_data", "ab_full.csv"))
trait_imputed <- read_csv(here("1_raw_input_data", "Imputed_data_inc_NIS.csv"))
trait <- read_csv(here("1_raw_input_data", "trait_data_cleaned.csv"))
commercial_full <- read_csv(here("1_raw_input_data", "missing_commercial_full.csv"))

```



### (1) Add temperature data 

Add from the updated `env_var` which include the new Lisandro data as well as mean values for 9 location that bio-oracle did not have information on (done based on the other location in the site)

```{r}
col_order<-colnames(medata)

# remove env old column

medata<-medata %>% select(-tmean,-sal_mean,-pp_mean)

# join with the new env data

medata<-left_join(medata,env_var)

# set back to the same column order

medata<-medata %>% select(all_of(col_order))

# clean the r environment

rm(env_var)
rm(col_order)

```

### (2) Add a and b coeffitents and biomass calculation


```{r include=FALSE}

# keep only metadate and abundances

medata<-medata %>% select(1:24) 

# Add a & b and native or invasive data

medata<- left_join(medata,ab_full)

# fix weird observation

medata$sp.length[medata$species =="Symphodus.tinca" & medata$sp.length == 120 & is.na(medata$sp.length) == F]<-12

medata$sp.n[medata$species =="Balistes.capriscus" & medata$sp.n == 2000]<-2

medata<-medata %>% mutate(biomass_single = a*sp.length^b) # in gram


# biomass * abundance

medata<-medata %>% mutate(biomass = biomass_single*sp.n) # in gram

medata$biomass_single<-NULL

# remove obs of 0 in abundance - some remenents from the access files where insted of delete you had to fill 0 sp.n


medata<-medata %>% filter(sp.n>0)

# remove obs of 0 in lenght or NA


medata<-medata %>% filter(sp.length>0) %>% drop_na(sp.length)

# convert the species names format to match the trait data

medata<-medata %>% mutate(species = gsub("\\.", " ",species))

# create the site-level identifier 

medata<-medata %>% mutate(site_id = paste(Site,enforcement,tmean,sep="_"),.after = Site)

# remove enforcement = 1 from data


medata<-medata %>% filter(enforcement!=1)

# NA in exotic column = Family level id...remove (mostly Clupeidae)

medata<-medata %>% drop_na(exotic)

rm(ab_full)

```

Update some names to match the trait data

```{r include=FALSE}

medata$species[medata$species=="Liza aurata"]<-"Chelon auratus"

medata$species[medata$species=="Oblada melanura"]<-"Oblada melanurus"

medata$species[medata$species=="Pempheris vanicolensis"]<-"Pempheris rhomboidea"

medata$species[medata$species=="Symphodus melanocercus"]<-"Centrolabrus melanocercus"

medata$species[medata$species=="Taeniura grabata"]<-"Taeniurops grabatus"

```

### (3) calculate invasive and native biomass per site & enforcment


calculate transects/site native,invasive and total biomass

first calculate as the transect level and than use the mean per site

```{r}


# calculate the total biomass of invasive and natives for each transect

transect_biomass <- medata %>%
  group_by(site_id,Site,tmean,enforcement,unique_trans_id,exotic) %>% 
  summarise(transect_biomass = sum(biomass,na.rm = T)) # units are g


# convert to wide format

transect_biomass<-spread(transect_biomass,exotic,transect_biomass)


# convert na in invasive biomass to 0 biomass


transect_biomass$Invasive[is.na(transect_biomass$Invasive)]<-0


# add total biomass


transect_biomass$total_biomass<-(transect_biomass$Invasive+transect_biomass$Native)



# calculate the mean biomass per site (separated to native and invasive)

site_biomass<-transect_biomass %>% 
  group_by(site_id,Site,tmean,enforcement) %>% 
  summarise(mean_invasive_biomass = mean(Invasive,na.rm = T),
            mean_native_biomass = mean(Native,na.rm = T),
            mean_total_biomass = mean(total_biomass,na.rm = T))
            

# convert enforcement to factor

site_biomass$enforcement<-as.factor(site_biomass$enforcement)


```

### (4) arrange trait data

check which species have traits


```{r include=FALSE}


#lets see is the species match

species<- medata %>% distinct(species)

# change traits data colnames

trait_imputed<-trait_imputed %>% rename("species" = "fishbase_name")

species_traits<-left_join(species,trait_imputed)

# add species that are not in the inputed data but have data in the older trait data

not_in_imp<-species_traits %>% filter(is.na(Home_Range)) %>% select(species)

trait<-trait %>% rename("species" = "fishbase_name")

not_in_imp <- left_join(not_in_imp,trait)

```

keep only relevant traits

```{r include=FALSE}

species_traits <- species_traits %>% 
    drop_na(Family) %>%  # drop the species that dot have data 
  select(species,Home_Range,Activity,Schooling,Water_Level,Diet)

# same for the species that are not in the imputed data

not_in_imp <- not_in_imp %>% 
  drop_na(Family) %>%  # drop the species that dot have data 
  select(species,Home_Range,Activity,Schooling,Water_Level,Diet)


# unite

species_traits<-bind_rows(species_traits,not_in_imp)

rm(not_in_imp,trait)

```

get the species max size to use for species size

```{r}
sp_size <- medata %>%
  group_by(species) %>%
  summarise(sp_size = max(sp.length))

# fix weird values

# we only saw juvenile Sphyraena Sphyraena- change to adult size according to common size in fishbase

sp_size$sp_size[sp_size$species=="Sphyraena sphyraena"]<-60

```

add species size based on the data

```{r}
species_traits<-left_join(species_traits,sp_size)

rm(sp_size)
```


```{r include=FALSE}

# make sure there is no duplicated rows

species_traits <- species_traits %>% distinct(species,Home_Range,Activity,Schooling,Water_Level,Diet,sp_size)

# count how many traits have NA for each species

species_traits$all_na<-rowSums(is.na(species_traits))

species_traits <- species_traits %>%
  mutate_at(2:6,factor) %>% # convert the character columns to Factors
  arrange(species) %>% # arrange the species alphabetically (so it would be similar to `my_data` order)
  filter(all_na<3) %>% # filter species with not enough traits data
  select(-all_na) # remove this non-trait columns

# add if exotic or native 

exotic_stat<-medata %>% distinct(species,exotic)

species_traits<-left_join(species_traits,exotic_stat)

```


```{r include=FALSE}

# order the Home range level

species_traits$Home_Range <- ordered(species_traits$Home_Range,levels = c("Sed","Mob","Vmob")) 

# order the Schooling levels

unique(species_traits$Schooling)

species_traits$Schooling<-droplevels(species_traits$Schooling)

species_traits$Schooling <-ordered(species_traits$Schooling,c("Sol","Pair","SmallG","MedG","LargeG"))


# order the water-level levels

species_traits$Water_Level <- ordered(species_traits$Water_Level,c("Bottom","Low","High"))

str(species_traits)



```

### (5) commercial value data


add species commercial data 

```{r}
species<-medata %>% distinct(species)

# join

commercial<-left_join(species,commercial_full)

# unite low to medium so ill have only 3 levels

commercial$Commercial_value<-ifelse(commercial$Commercial_value=="Low","Medium",commercial$Commercial_value)

# convert to factor

commercial$Commercial_value<-as.factor(commercial$Commercial_value)

# order the factor

commercial$Commercial_value<- ordered(commercial$Commercial_value,levels = c("None","Medium","High")) 

# later i need to revise my classification but now ill remove this column

commercial$source<-NULL

# unite with medata

medata_commer<-left_join(medata,commercial)

# add native or nis to commercial

native<-medata %>% distinct(species,exotic)

commercial<-left_join(commercial,native)

rm(native,commercial_full)

```

calculate the total biomass of invasive and natives for each commercial value for each transect


```{r}

transect_biomass_com <- medata_commer %>%
  group_by(site_id,Site,tmean,enforcement,unique_trans_id,exotic,Commercial_value) %>% 
  summarise(transect_biomass = sum(biomass,na.rm = T))


# calculate the mean biomass per site (separated to native and invasive)

site_biomass_com<-transect_biomass_com %>% 
  group_by(site_id,Site,tmean,enforcement,exotic,Commercial_value) %>% 
  summarise(mean_biomass = mean(transect_biomass,na.rm = T))


# convert enforcement to factor

site_biomass_com$protection <- ifelse(site_biomass_com$enforcement>0,"Protected","Not protected")


site_biomass_com$enforcement<-as.factor(site_biomass_com$enforcement)



```


```{r}

# Im keeping only natives communities biomass

biomass_commer <- site_biomass_com  %>% filter(exotic=="Native")

# remove this column

biomass_commer$exotic<-NULL

#convert to factor

biomass_commer$protection<-as.factor(biomass_commer$protection)

# add invasive species from the site_biomass (not separate to commercial value)

inv_bio<-site_biomass %>% select(site_id,mean_invasive_biomass)

# add invasive biomass (not divided by commercial value)

biomass_commer<-left_join(biomass_commer,inv_bio)


# change column names to match to diversity table (created with i-next in the next scripts)

colnames(biomass_commer)<-c("Assemblage","Site","tmean" ,"enforcement","Commercial_value",
"Native_biomass","protection","Invasive_biomass")

```

### (6) convert all biomass estimations to kg in the data that are saved for letter use

```{r}
# convert weights to kg

medata <- medata %>% mutate(biomass = biomass/1000)

site_biomass<-site_biomass %>%  mutate(mean_invasive_biomass = mean_invasive_biomass/1000,
                                       mean_native_biomass = mean_native_biomass/1000,
                                       mean_total_biomass = mean_total_biomass/1000)

biomass_commer<-biomass_commer %>% mutate(Native_biomass = Native_biomass/1000,Invasive_biomass=Invasive_biomass/1000)

```


### (7) save data



```{r}


#save(medata, file = here("2_data_for_analyses", "medata_multistressor.rdata")) # medata clean for this project

#save(species_traits, file = here("2_data_for_analyses", "species_traits_multistressor.rdata")) #t rait data clean for this project

#save(site_biomass, file = here("2_data_for_analyses", "site_biomass.rdata")) # table of the native and total biomass per site

#save(biomass_commer, file = here("2_data_for_analyses", "site_biomass_commercial.rdata"))  #  native and invasive biomass per site & commercial value 

#save(commercial, file = here("2_data_for_analyses", "final_commercial_list.rdata")) # final list of  species commercial values


```


