---
title: "Clean 4 - i next null models"
author: "TG"
date: "2026-05-14"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
# In this script:

i-next calculation to rarefy TD and FD diversity for native communities

*Input data:*


* `medata_multistressor.rdata` - medata clean for this project

* `species_traits_multistressor.rdata` - traits data clean for this project with imputations

* `site_biomass.rdata` - table of the native and total biomass per site

* input files are at `2_data_for_analyses` folder

*output data:*

* `diversity_table_natives.rdata` - rarefied TD and FD diversities of native communities

* `results_FD_native_only.Rdata` - since it computational heavy this sub-table is also saved for easier down stream code fixes 

* outputs saved in `3_diversity tables` folder

### Load packeges

```{r message=FALSE, warning=FALSE}

library(iNEXT.3D)
library(tidyverse)
library(vegan)
library(plotrix)
library(sjPlot)
library(sjmisc)
library(gridExtra)
library(ggeffects)
library(vegan)
library(plotly)
library(FD)
library(mFD)
library(funrar)
library(here)

select<-dplyr::select

```

### load data

```{r}

load(here("2_data_for_analyses", "medata_multistressor.rdata"))

load(here("2_data_for_analyses", "species_traits_multistressor.rdata"))

load(here("2_data_for_analyses", "site_biomass.rdata"))

```


### (1) count how many invasive individuals and species i have in each transect (how many native species fillers I need here)

```{r}
invasive_count <-medata %>% filter(tmean>19) %>%  
                      group_by(unique_trans_id,exotic) %>% 
                      summarise(invasive_abundance = sum(sp.n),invasive_species = n_distinct(species)) %>% 
                      filter(exotic == "Invasive")

invasive_count$exotic<-NULL


  
```

### (2) Build a species pool

based on the natives that are found in temp > 19 degree

```{r}

natives_pool <-medata %>% filter(exotic == "Native") %>%  filter(tmean >19) %>% group_by(species) %>% summarise(total_abundance = sum(sp.n))

```



### (3) remove invasive from the data and unite with the filling instructions



```{r}
medata_19<-medata %>% filter(tmean>19)

medata_19 <- medata_19 %>% filter(exotic == "Native")

medata_19<-left_join(medata_19,invasive_count)
```



```{r}

medata_native<-medata %>% filter(exotic == "Native")



medata_native<-medata_native %>% group_by(across(1:23)) %>% summarise(amount = sum(sp.n)) %>% ungroup() # sum the individuals from the same species in each transect

medata_native_wide<-spread(medata_native,species,amount,fill = 0)


```
