---
title: "Clean 5 - relative importance"
author: "TG"
date: "2026-05-15"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

# In this script: 

create the relative importance plots for all diversity and biomass facets (including differnt commertial values)


*Input data:*

* `diversity_table_natives.rdata` - rarefied TD and FD diversities of native communities

* `medata_multistressor.rdata` - medata clean for this project

* `site_biomass_commercial.rdata`- biomass divided by natives commercial value

* `final_commercial_list.rdata`- the final list of the species commercial values


*Output data:*

* all relative importance plots

* ternary plots + interactions for diversity & biomass

* ternary plots for different commercial values 



### load packages

```{r}
library(tidyverse)
library(relaimpo)
library(patchwork)
library(ggiraphExtra)
library(ggeffects)
library(ggtern)
select<-dplyr::select

rela_plots<-"4_relative importance"

```



### load data

```{r}

load(here("2_data_for_analyses", "medata_multistressor.rdata"))
load(here("2_data_for_analyses", "site_biomass_commercial.rdata"))
load(here("2_data_for_analyses", "final_commercial_list.rdata"))

load(here("3_diversity tables", "diversity_table_natives.rdata"))

```



# Create lm

diversity models


```{r}

# diversity

TD_richness<-lm(TD_richness ~ tmean + protection + Invasive_biomass,data = diversity_table)

TD_shannon<-lm(TD_shannon ~ tmean + protection + Invasive_biomass,data = diversity_table)

TD_simpson<-lm(log(TD_simpson) ~ tmean + protection + Invasive_biomass,data = diversity_table)

FD_richness<-lm(FD_richness ~ tmean + protection + Invasive_biomass,data = diversity_table)

FD_shannon<-lm(FD_shannon ~ tmean + protection + Invasive_biomass,data = diversity_table)

FD_simpson<-lm(FD_simpson ~ tmean + protection + Invasive_biomass,data = diversity_table)


```

biomass model

```{r}
# split the commercial data by value

high <- biomass_commer %>% filter(Commercial_value=="High")
medium <- biomass_commer %>% filter(Commercial_value=="Medium")
none <- biomass_commer %>% filter(Commercial_value=="None")

# check if normal distributions

hist(high$Native_biomass)
hist(medium$Native_biomass)
hist(none$Native_biomass)

# nope - convert to log...

hist(log(high$Native_biomass))
hist(log(medium$Native_biomass))
hist(log(none$Native_biomass))

```

lm models

```{r}

# biomass (total)

Biomass<-lm(log(Native_biomass) ~ tmean + protection + Invasive_biomass,data = diversity_table)

# set of linear models 

bio_high<-lm(log(Native_biomass)~ tmean + protection + Invasive_biomass,data = high)

bio_medium<-lm(log(Native_biomass)~ tmean + protection + Invasive_biomass,data = medium)

bio_none<-lm(log(Native_biomass)~ tmean + protection + Invasive_biomass,data = none)

```


# Relative importance

a function that calculate, plot and save the relative importance for each y

`boot.relimp()` - conducts the bootstrap

`booteval.relimp()` - gets the confidence intervals

it also provide the relative in data fit to plot ternary plots

```{r}


plot_my_rela <- function(model, response_name, plot_title = response_name,
                         save_plot = FALSE, upperl = 1, rela_plots = NULL) {
  
  # bootstrap relative importance
  
  rl_boot <- boot.relimp(model, b = 1000, rela = TRUE)
  ci <- booteval.relimp(rl_boot)
  
  print(ci)
  
  # data for bar plot
 
  data <- as.data.frame(ci@mark)
  colnames(data) <- c("var_explained", "sign", "lower_ci", "upper_ci")
  
  data <- data %>%
    mutate(
      var_explained = as.numeric(var_explained),
      lower_ci = as.numeric(lower_ci),
      upper_ci = as.numeric(upper_ci),
      variable = rownames(data),
      variable = case_when(
        grepl("tmean", variable) ~ "Temperature",
        grepl("protection", variable) ~ "Protection",
        grepl("Invasive_biomass", variable) ~ "NIS"
      ),
      variable = ordered(variable, c("Temperature", "Protection", "NIS"))
    )
  
  total_var <- ci@R2
  
  # data for ternary plot
  rela_out <- data %>%
    select(variable, var_explained) %>%
    pivot_wider(names_from = variable, values_from = var_explained) %>%
    mutate(response = response_name) %>%
    select(response, Temperature, Protection, NIS)
  
  # bar plot
  plot <- ggplot(data = data, aes(x = variable, y = var_explained, fill = variable)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    ylab("Variance explained") +
    labs(
      title = plot_title,
      subtitle = bquote(R^2 == .(round((total_var * 100), 2)) * "%")
    ) +
    ylim(0, upperl) +
    theme(
      legend.position = "none",
      axis.title.x = element_blank(),
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 24),
      plot.title = element_text(size = 26, hjust = 0.5),
      plot.subtitle = element_text(size = 18, hjust = 0.5)
    ) +
    scale_fill_manual(values = c("#364D8B", "#03718A", "#30A3D1"))
  
  
  if (save_plot) {
    plot_name <- paste0("Relative importance - ", plot_title, ".pdf")
    ggsave(
      plot = plot,
      filename = plot_name,
      width = 10,
      height = 8,
      dpi = 300,
      path = rela_plots
    )
  }
  
  return(list(
    plot = plot,
    rela = rela_out,
    ci_data = data,
    R2 = total_var
  ))
}

```


run over diversity and biomass indeces

```{r}

td_richness_rela <- plot_my_rela(model = TD_richness, response_name = "TD richness", upperl = 0.7, save_plot = FALSE)

td_shannon_rela <- plot_my_rela(model = TD_shannon, response_name = "TD shannon", upperl = 1, save_plot = FALSE)

td_simpson_rela <- plot_my_rela(model = TD_simpson, response_name = "TD simpson", upperl = 1, save_plot = FALSE)

fd_richness_rela <- plot_my_rela(model = FD_richness, response_name = "FD richness", upperl = 0.7, save_plot = FALSE)

fd_shannon_rela <- plot_my_rela(model = FD_shannon, response_name = "FD shannon", upperl = 1, save_plot = FALSE)

fd_simpson_rela <- plot_my_rela(model = FD_simpson, response_name = "FD simpson", upperl = 1, save_plot = FALSE)

biomass_all_rela <- plot_my_rela(model = Biomass, response_name = "Native biomass", upperl = 1, save_plot = FALSE)

biomass_high_rela <- plot_my_rela(model = bio_high, response_name = "High commercial value", upperl = 1, save_plot = FALSE)
biomass_medium_rela <- plot_my_rela(model = bio_medium, response_name = "Medium commercial value", upperl = 1, save_plot = FALSE)
biomass_none_rela <- plot_my_rela(model = bio_none, response_name = "No commercial value", upperl = 1, save_plot = FALSE)

```


## ternary plots

all the data wrangling needed

```{r}


rela_diversity_df <- bind_rows(
  td_richness_rela$rela,
  td_shannon_rela$rela,
  td_simpson_rela$rela,
  fd_richness_rela$rela,
  fd_shannon_rela$rela,
  fd_simpson_rela$rela,
  biomass_all_rela$rela) %>% 
                        mutate(facet = case_when(
                                      grepl("^TD", response) ~ "Taxonomic",
                                      grepl("^FD", response) ~ "Functional",
                                      response == "Native biomass" ~ "Biomass"),
                               index = case_when(
                                      grepl("richness", response, ignore.case = TRUE) ~ "Richness",
                                      grepl("shannon", response, ignore.case = TRUE) ~ "Shannon",
                                      grepl("simpson", response, ignore.case = TRUE) ~ "Simpson",
                                      response == "Native biomass" ~ "Biomass"))

div_colors <- c(
  "TD richness" = "#08306B",
  "TD shannon" = "#4292C6",
  "TD simpson" = "#9ECAE1",
  "FD richness" = "#67000D",
  "FD shannon" = "#EF3B2C",
  "FD simpson" = "#FCAE91",
  "Native biomass" ="#41AE76"
)

```

Ternary plot

```{r}
div_ternary <- ggtern(rela_diversity_df,
                             aes(x = Temperature, y = Protection, z = NIS)) +
  geom_point(aes(color = response, shape = response), size = 3.2, alpha = 0.95) +
  labs(x = "Temperature", y = "Protection", z = "NIS", color = NULL, shape = NULL) +
  scale_color_manual(values = div_colors) +
  scale_shape_manual(values = c("TD richness" = 16,
                                "TD shannon" = 16,
                                "TD simpson" = 16,
                                "FD richness" = 17,
                                "FD shannon" = 17,
                                "FD simpson" = 17,
                                "Native biomass" = 15)) +
  theme_bw(base_size = 8) +
  theme_showarrows() +
  theme(legend.position = "top",
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        tern.axis.title.T = element_text(size = 10),
        tern.axis.title.L = element_text(size = 10),
        tern.axis.title.R = element_text(size = 10),
        tern.axis.text.T = element_text(size = 7),
        tern.axis.text.L = element_text(size = 7),
        tern.axis.text.R = element_text(size = 7),
        tern.axis.arrow.text.T = element_text(size = 7),
        tern.axis.arrow.text.L = element_text(size = 7),
        tern.axis.arrow.text.R = element_text(size = 7),
        tern.panel.grid.major = element_line(color = "grey88", linewidth = 0.25),
        plot.margin = margin(1, 1, 1, 1))

div_ternary

```



```{r}
# ggsave(here("4_relative importance", "Ternary_relative_importance_diversity.pdf"),
#        plot = div_ternary, width = 8.5, height = 8.5, units = "cm")
```



### relative importance plots for biomass models

```{r}

rela_biomass_df <- bind_rows(
  biomass_high_rela$rela,
  biomass_medium_rela$rela,
  biomass_none_rela$rela)


rela_biomass_df <- rela_biomass_df %>%
  mutate(response_short = recode(response,
                                 "High commercial value" = "High",
                                 "Medium commercial value" = "Medium",
                                 "No commercial value" = "No value"))

bio_colors <- c("High" = "#00798C",
                "Medium" = "#70A288",
                "No value" = "#D9ED92")

```



```{r}



bio_ternary <- ggtern(rela_biomass_df,aes(x = NIS, y = Temperature, z = Protection)) +
  geom_point(aes(color = response_short, size = response_short), shape = 16, alpha = 0.95) +
  labs(x = "NIS", y = "Temperature", z = "Protection", color = NULL, size = NULL) +
  scale_color_manual(values = bio_colors) +
  scale_size_manual(values = c("Total" = 4.2,"High" = 3.0,"Medium" = 3.0,"No value" = 3.0)) +
  theme_bw(base_size = 8) +
  theme_showarrows() +
  theme(legend.position = "right",
        legend.text = element_text(size = 7),
        legend.key.size = unit(0.35, "cm"),
        tern.axis.title.T = element_text(size = 9),
        tern.axis.title.L = element_text(size = 9),
        tern.axis.title.R = element_text(size = 9),
        tern.axis.text.T = element_text(size = 6),
        tern.axis.text.L = element_text(size = 6),
        tern.axis.text.R = element_text(size = 6),
        tern.axis.arrow.text.T = element_text(size = 6),
        tern.axis.arrow.text.L = element_text(size = 6),
        tern.axis.arrow.text.R = element_text(size = 6),
        tern.panel.grid.major = element_line(color = "grey90", linewidth = 0.2),
        plot.margin = margin(3, 10, 3, 3))

bio_ternary

# ggsave(
#   filename = here("4_relative importance", "Ternary_relative_importance_biomass.pdf"),
#   plot = bio_ternary,
#   width = 8.5,
#   height = 8.5,
#   units = "cm"
# )
```


# Interaction model 

create interaction model and prepare for ploting for each Y variable

temp * protection 

### TD Richness

```{r}

i_TD_richness<-lm(TD_richness ~ tmean *  protection ,data = diversity_table)

summary(i_TD_richness)

p_i_TD_richness <- ggpredict(i_TD_richness, 
                       terms = c('tmean[16:24 by=0.5]',
                                 "protection"))

colnames(p_i_TD_richness)<-c("Temperature","predicted","std.error","conf.low","conf.high","Protection" )

```


### TD shannon


```{r}
i_TD_shannon<-lm(TD_shannon ~ tmean *  protection ,data = diversity_table)

summary(i_TD_shannon)

p_i_TD_shannon <- ggpredict(i_TD_shannon, 
                       terms = c('tmean[16:24 by=0.5]',
                                 "protection"))

colnames(p_i_TD_shannon)<-c("Temperature","predicted","std.error","conf.low","conf.high","Protection" )


```


### TD simpson

```{r}
i_TD_simpson<-lm(log(TD_simpson) ~ tmean *  protection ,data = diversity_table)

summary(i_TD_simpson)

p_i_TD_simpson <- ggpredict(i_TD_simpson, 
                       terms = c('tmean[16:24 by=0.5]',
                                 "protection"))

colnames(p_i_TD_simpson)<-c("Temperature","predicted","std.error","conf.low","conf.high","Protection" )



```
### FD richness

```{r}
i_FD_richness<-lm(FD_richness ~ tmean *  protection ,data = diversity_table)

summary(i_FD_richness)

p_i_FD_richness <- ggpredict(i_FD_richness, 
                       terms = c('tmean[16:24 by=0.5]',
                                 "protection"))

colnames(p_i_FD_richness)<-c("Temperature","predicted","std.error","conf.low","conf.high","Protection" )
```


### FD shannon

```{r}
i_FD_shannon<-lm(FD_shannon ~ tmean *  protection ,data = diversity_table)

summary(i_FD_shannon)

p_i_FD_shannon <- ggpredict(i_FD_shannon, 
                       terms = c('tmean[16:24 by=0.5]',
                                 "protection"))

colnames(p_i_FD_shannon)<-c("Temperature","predicted","std.error","conf.low","conf.high","Protection" )
```
### FD simpson


```{r}
i_FD_simpson<-lm(FD_simpson ~ tmean *  protection ,data = diversity_table)

summary(i_FD_simpson)

p_i_FD_simpson <- ggpredict(i_FD_simpson, 
                       terms = c('tmean[16:24 by=0.5]',
                                 "protection"))

colnames(p_i_FD_simpson)<-c("Temperature","predicted","std.error","conf.low","conf.high","Protection" )



```


```{r}
i_biomass<-lm(Native_biomass ~ tmean *  protection ,data = diversity_table)

summary(i_biomass)

p_i_biomass <- ggpredict(i_biomass, 
                       terms = c('tmean[16:24 by=0.5]',
                                 "protection"))

colnames(p_i_biomass)<-c("Temperature","predicted","std.error","conf.low","conf.high","Protection" )
```


### interaction plot function & theme


```{r}
interaction_theme <- theme_classic(base_size = 8) +
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size = 9),
        legend.position = c(0.68, 0.88),
        legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.75), color = NA),
        legend.text = element_text(size = 8),
        legend.title = element_blank(),
        legend.key.size = unit(0.35, "cm"),
        plot.margin = margin(3, 3, 3, 3))

make_interaction_plot <- function(pred_data, raw_y, ylab_text, col_hex) {
  ggplot(data = pred_data) +
    geom_ribbon(aes(x = Temperature, ymin = conf.low, ymax = conf.high, fill = Protection), alpha = 0.2) +
    geom_line(aes(x = Temperature, y = predicted, color = Protection), linewidth = 0.7) +
    geom_point(data = diversity_table, aes(x = tmean, y = {{ raw_y }}, color = protection), alpha = 0.25, size = 0.8) +
    xlab("Temperature (°C)") +
    ylab(ylab_text) +
    scale_color_manual(values = c("grey60", col_hex)) +
    scale_fill_manual(values = c("grey60", col_hex)) +
    guides(fill = "none") +
    interaction_theme
}

```


plot interaction plots

```{r}


p_td_rich <- make_interaction_plot(p_i_TD_richness, TD_richness, "TD richness", "#08306B")
p_td_shan <- make_interaction_plot(p_i_TD_shannon, TD_shannon, "TD Shannon", "#4292C6")
p_td_simp <- make_interaction_plot(p_i_TD_simpson, TD_simpson, "TD Simpson", "#9ECAE1")

p_fd_rich <- make_interaction_plot(p_i_FD_richness, FD_richness, "FD richness", "#67000D")
p_fd_shan <- make_interaction_plot(p_i_FD_shannon, FD_shannon, "FD Shannon", "#EF3B2C")
p_fd_simp <- make_interaction_plot(p_i_FD_simpson, FD_simpson, "FD Simpson", "#FCAE91")

p_biomass <- make_interaction_plot(p_i_biomass, Native_biomass, "Native biomass", "#41AE76")

# combined layout


layout_design <- "
ABC
DBE
FGH
"

combined_diversity_figure <- wrap_plots(
  A = p_td_rich,
  B = div_ternary,
  C = p_fd_rich,
  D = p_td_shan,
  E = p_fd_shan,
  F = p_td_simp,
  G = p_biomass,
  H = p_fd_simp,
  design = layout_design) + 
  plot_layout(widths = c(1, 1.15, 1),
              heights = c(1, 1, 1))

combined_diversity_figure

# ggsave(filename = here("4_relative importance",
#                        "Combined_interactions_ternary_diversity.pdf"),
#        plot = combined_diversity_figure,
#        width = 24,
#        height = 18,
#        units = "cm")
```


### create the diversity multuple plot - old style


```{r}
p_a <- td_richness_rela[["plot"]]
p_c <- td_shannon_rela[["plot"]]
p_e <- td_simpson_rela[["plot"]]

p_b <- fd_richness_rela[["plot"]]
p_d <- fd_shannon_rela[["plot"]]
p_f <- fd_simpson_rela[["plot"]]

                       # 3 rows, 2 columns
combined_plot <- (p_a | p_b) /
                 (p_c | p_d) /
                 (p_e | p_f)
# View it
combined_plot

# ggsave("4_relative importance/All_relative_importance_panels.pdf", plot = combined_plot,width = 14, height = 12, dpi = 300)
# ggsave("4_relative importance/All_relative_importance_panels.png", plot = combined_plot,width = 14, height = 12, dpi = 300)


rm(p_a,p_b,p_c,p_d,p_e,p_f)
```

save table of all relative info

```{r}
rela_info<-rela_diversity_df %>% select(response,Temperature,Protection,NIS)
rela_bio<- rela_biomass_df %>% select(response,Temperature,Protection,NIS)

rela_info<-bind_rows(rela_info,rela_bio)

r2_info <- tibble(
  response = c("TD richness", "TD shannon", "TD simpson",
               "FD richness", "FD shannon", "FD simpson",
               "Native biomass", "High commercial value",
               "Medium commercial value", "No commercial value"),
  R2 = c(td_richness_rela$R2,
         td_shannon_rela$R2,
         td_simpson_rela$R2,
         fd_richness_rela$R2,
         fd_shannon_rela$R2,
         fd_simpson_rela$R2,
         biomass_all_rela$R2,
         biomass_high_rela$R2,
         biomass_medium_rela$R2,
         biomass_none_rela$R2))

rela_info <- rela_info %>%left_join(r2_info, by = "response")%>%
  mutate(across(c(Temperature, Protection, NIS, R2), ~round(.x * 100, 1)))

write_csv(rela_info,
          here("4_relative importance",
               "Table_relative_importance.csv"))

```




