---
title: "Clean 4 - i next null models"
author: "TG"
date: "2026-05-14"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
# In this script:

i-next calculation to rarefy TD and FD diversity for native communities

*Input data:*


* `medata_multistressor.rdata` - medata clean for this project

* `species_traits_multistressor.rdata` - traits data clean for this project with imputations

* `site_biomass.rdata` - table of the native and total biomass per site

* input files are at `2_data_for_analyses` folder

*output data:*

* `diversity_table_natives.rdata` - rarefied TD and FD diversities of native communities

* `results_FD_native_only.Rdata` - since it computational heavy this sub-table is also saved for easier down stream code fixes 

* outputs saved in `3_diversity tables` folder

### Load packeges

```{r message=FALSE, warning=FALSE}

library(iNEXT.3D)
library(tidyverse)
library(vegan)
library(plotrix)
library(sjPlot)
library(sjmisc)
library(gridExtra)
library(ggeffects)
library(vegan)
library(plotly)
library(FD)
library(mFD)
library(funrar)
library(here)
library(pbapply)

select<-dplyr::select

c_min <- 0.9540893
```

### load data

```{r}

load(here("2_data_for_analyses", "medata_multistressor.rdata"))

load(here("2_data_for_analyses", "species_traits_multistressor.rdata"))

load(here("2_data_for_analyses", "site_biomass.rdata"))

```


functions:


```{r}

draw_native_species <- function(n) {
  sample(
    x = natives_pool$species,
    size = n,
    replace = TRUE,
    prob = natives_pool$total_abundance
  )
}


create_null_community <- function() {
  
  null_invasives <- invasive_count %>%
    rowwise() %>%
    mutate(species = list(draw_native_species(invasive_abundance))) %>%
    unnest(species) %>%
    mutate(abundance = 1) %>%
    group_by(unique_trans_id, species) %>%
    summarise(sp.n = sum(abundance), .groups = "drop")
  
  native_replacement <- left_join(null_invasives, transect_mdata, by = "unique_trans_id")
  
  native_replacement <- native_replacement %>%
    relocate(unique_trans_id, .after = trans) %>%
    relocate(species, .after = pp_mean) %>%
    relocate(sp.n, .after = species)
  
  medata_only_natives <- medata_19 %>%
    select(-sp.length, -biomass, -a, -b) %>%
    filter(exotic != "Invasive") %>%
    select(-exotic) %>%
    bind_rows(native_replacement) %>%
    group_by(across(1:23)) %>%
    summarise(sp.n = sum(sp.n), .groups = "drop")
  
  medata_natives_wide<-spread(medata_only_natives,species,sp.n,fill = 0)
  
  site_id<-medata_natives_wide %>%
  mutate(site_id = paste(Site,enforcement,tmean,sep="_"),.after = Site) %>% drop_na(tmean)

# keep just the id_column and the species and unite the counts 

site_id<- site_id %>%
  select(site_id,23:ncol(site_id)) %>%
  group_by(site_id) %>%
  summarise(across(.fns = sum), .groups = "drop")

# move the id_column to be the rowname

site_id<-site_id %>% column_to_rownames(var = 'site_id')

# transpose so the species are in the rows and site in column

site_id<- t(site_id)

  
    return(site_id)
}


process_td_results <- function(results_td) {
  
  td.diversity <- left_join(results_td, meta.data)
  
  td.diversity$enforcement <- as.factor(td.diversity$enforcement)
 
   diversity_table <- td.diversity %>%
  select(c(11:14, SC, Assemblage, Order.q, qTD)) %>%
  mutate(Order.q = case_when(
  Order.q == 0 ~ "TD_richness",
  Order.q == 1 ~ "TD_shannon",
  Order.q == 2 ~ "TD_simpson")) %>%
  spread(key = Order.q, value = qTD) %>%

  left_join(site_biomass)

  return(diversity_table)

}

make_fd_matrix <- function(site_id_data) {
  
# filter the data accordingly

site_id_fd<-as.data.frame(site_id_data) #species in rows,sites in columns

site_id_fd <-site_id_fd %>% filter(row.names(site_id_fd) %in% keep)

# convert back to data.matrix

site_id_fd<-data.matrix(site_id_fd)

return(site_id_fd)

}

```


### ill start with filtering only transect of >19

```{r}

medata_19<-medata %>% filter(tmean>=19)

```

# null model based on abundance

### (1) count how many invasive individuals and species i have in each transect (how many native species fillers I need here)

```{r}
invasive_count <-medata_19 %>% group_by(unique_trans_id,exotic) %>% 
                      summarise(invasive_abundance = sum(sp.n))%>% 
                      filter(exotic == "Invasive")

invasive_count$exotic<-NULL


```

### (2) Build a species pool

based on the natives that are found in temp > 19 degree

```{r}

natives_pool <-medata_19 %>% filter(exotic == "Native") %>% group_by(species) %>% summarise(total_abundance = sum(sp.n))

```


```{r}
transect_mdata<-medata_19 %>% select(1:22) %>% distinct()
```


keep trait for natives only (>19)


```{r}
species_traits_native<-species_traits %>% filter(exotic=="Native")

species_traits_native$exotic<- NULL
```


convert to gower distance matrix and save in "data.matrix" format


```{r include=FALSE}



# add the species as row names

my_traits <- species_traits_native %>% filter(species %in% natives_pool$species)%>% column_to_rownames(var="species") 


traits_gower_dist <- gowdis(my_traits)
traits_gower_matrix<-data.matrix(traits_gower_dist)

keep <- rownames(traits_gower_matrix)

```

# create the 100 null communities

```{r}

set.seed(1)

null_medata_list <- replicate(100,create_null_community(), simplify = FALSE)

identical(null_medata_list[[1]], null_medata_list[[2]]) # should be false


```


```{r}


results_td_list <- lapply(null_medata_list,function(x) estimate3D(
      x,
      diversity = "TD",
      q = c(0, 1, 2),
      datatype = "abundance",
      base = "coverage",
      level = c_min))



# connect to the medata


meta.data<-transect_mdata %>% 
  mutate(Assemblage = paste(Site,enforcement,tmean,sep="_"),.after = Site) %>%  
  drop_na(tmean) %>%
  select(Assemblage,Basin,Site,enforcement,protection,tmean) %>%
  distinct(Assemblage,.keep_all = T)



null_diversity_tables <- lapply(results_td_list, process_td_results)

#save(null_diversity_tables, file = here("3_diversity tables", "results_td_null_list.Rdata"))

```


```{r}
results_fd_test <- pblapply(null_medata_list[1:2], function(x) {

  site_id_fd <- make_fd_matrix(x)

  estimate3D(site_id_fd,
             diversity = "FD",
             q = c(0, 1, 2),
             FDdistM = traits_gower_dist,
             datatype = "abundance",
             base = "coverage",
             level = c_min)})
```



```{r}
results_fd_list <- pblapply(null_medata_list, function(x) {

  site_id_fd <- make_fd_matrix(x)

  estimate3D(site_id_fd,
             diversity = "FD",
             q = c(0, 1, 2),
             FDdistM = traits_gower_dist,
             datatype = "abundance",
             base = "coverage",
             level = c_min)})

```



```{r}
results_fd_list <- pblapply(null_medata_list, function(x) {

  site_id_fd <- make_fd_matrix(x)

  estimate3D(site_id_fd,
             diversity = "FD",
             q = c(0, 1, 2),
             FDdistM = traits_gower_dist,
             datatype = "abundance",
             base = "coverage",
             level = c_min)})

```



