---
title: "Clean 4 - i next null models"
author: "TG"
date: "2026-05-14"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
# In this script:

This script creates an abundance-based null model in which invasive individuals in warm-water transects (tmean ≥ 19°C) are replaced by randomly drawn native species. Native replacement species are sampled with replacement from the native species pool, weighted by their total abundance, so common native species are more likely to be selected. The script then calculates rarefied taxonomic diversity (TD) and functional diversity (FD) for each null community and combines the resulting diversity metrics with site-level metadata and biomass data.

*Input data:*

* `medata_multistressor.rdata` - cleaned Mediterranean fish transect data for this project

* `species_traits_multistressor.rdata` - cleaned and imputed species trait data

* `site_biomass.rdata` - site-level native, invasive, and total biomass estimates

* input files are located in the `2_data_for_analyses` folder

*Output data:*

* `results_td_list.Rdata` - rarefied TD estimates for all null communities

* `results_fd_null_list.Rdata` - rarefied FD estimates for all null communities; saved separately because this step is computationally intensive

* `null_abundance_diversity_df.Rdata` - final combined table of TD, FD, biomass, and metadata for all null iterations

* outputs are saved in the `3_diversity tables` folder



### Load packeges

```{r message=FALSE, warning=FALSE}

library(iNEXT.3D)
library(tidyverse)
library(vegan)
library(plotrix)
library(sjPlot)
library(sjmisc)
library(gridExtra)
library(ggeffects)
library(vegan)
library(plotly)
library(FD)
library(mFD)
library(funrar)
library(here)
library(pbapply)

select<-dplyr::select

c_min <- 0.9540893
```

### Load data

```{r}

load(here("2_data_for_analyses", "medata_multistressor.rdata"))

load(here("2_data_for_analyses", "species_traits_multistressor.rdata"))

load(here("2_data_for_analyses", "site_biomass.rdata"))

```


# Functions


```{r}

draw_native_species <- function(n) {
  sample(
    x = natives_pool$species,
    size = n,
    replace = TRUE,
    prob = natives_pool$total_abundance
  )
} 

create_null_community <- function() {
  
  null_invasives <- invasive_count %>%
    rowwise() %>%
    mutate(species = list(draw_native_species(invasive_abundance))) %>%
    unnest(species) %>%
    mutate(abundance = 1) %>%
    group_by(unique_trans_id, species) %>%
    summarise(sp.n = sum(abundance), .groups = "drop")
  
  native_replacement <- left_join(null_invasives, transect_mdata, by = "unique_trans_id")
  
  native_replacement <- native_replacement %>%
    relocate(unique_trans_id, .after = trans) %>%
    relocate(species, .after = pp_mean) %>%
    relocate(sp.n, .after = species)
  
  medata_only_natives <- medata_19 %>%
    select(-sp.length, -biomass, -a, -b) %>%
    filter(exotic != "Invasive") %>%
    select(-exotic) %>%
    bind_rows(native_replacement) %>%
    group_by(across(1:23)) %>%
    summarise(sp.n = sum(sp.n), .groups = "drop")
  
  medata_natives_wide<-spread(medata_only_natives,species,sp.n,fill = 0)
  
  site_id<-medata_natives_wide %>%
  mutate(site_id = paste(Site,enforcement,tmean,sep="_"),.after = Site) %>% drop_na(tmean)

# keep just the id_column and the species and unite the counts 

site_id<- site_id %>%
  select(site_id,23:ncol(site_id)) %>%
  group_by(site_id) %>%
  summarise(across(.fns = sum), .groups = "drop")

# move the id_column to be the rowname

site_id<-site_id %>% column_to_rownames(var = 'site_id')

# transpose so the species are in the rows and site in column

site_id<- t(site_id)

  
    return(site_id)
}

process_td_results <- function(results_td) {
  
  td.diversity <- left_join(results_td, meta.data)
  
  td.diversity$enforcement <- as.factor(td.diversity$enforcement)
 
  diversity_table <- td.diversity %>%
  select(c(11:14, SC, Assemblage, Order.q, qTD)) %>%
  mutate(Order.q = case_when(
  Order.q == 0 ~ "TD_richness",
  Order.q == 1 ~ "TD_shannon",
  Order.q == 2 ~ "TD_simpson")) %>%
  spread(key = Order.q, value = qTD) %>%

  left_join(site_biomass)

  return(diversity_table)

}

make_fd_matrix <- function(site_id_data) {
  
# filter the data accordingly

site_id_fd<-as.data.frame(site_id_data) #species in rows,sites in columns

site_id_fd <-site_id_fd %>% filter(row.names(site_id_fd) %in% keep)

# convert back to data.matrix

site_id_fd<-data.matrix(site_id_fd)

return(site_id_fd)

}

process_fd_results <- function(results_fd) {
  
  fd.diversity <- left_join(results_fd, meta.data)
  
  fd.diversity$enforcement <- as.factor(fd.diversity$enforcement)
 
  diversity_table_FD <- fd.diversity %>%
  select(c(11:14, SC, Assemblage, Order.q, qFD)) %>%
  mutate(Order.q = case_when(
  Order.q == 0 ~ "FD_richness",
  Order.q == 1 ~ "FD_shannon",
  Order.q == 2 ~ "FD_simpson")) %>%
  spread(key = Order.q, value = qFD) 

  return(diversity_table_FD)

}

```


# Data wrangling 

filtering only transect of >19

```{r}

medata_19<-medata %>% filter(tmean>=19)

```

# Null model based on abundance

(1) count how many invasive individuals and species i have in each transect (how many native species fillers I need here) 

```{r}
invasive_count <-medata_19 %>% group_by(unique_trans_id,exotic) %>% 
                      summarise(invasive_abundance = sum(sp.n))%>% 
                      filter(exotic == "Invasive")

invasive_count$exotic<-NULL


```

(2) Build a species pool

based on the natives that are found in temp > 19 degree

```{r}

natives_pool <-medata_19 %>% filter(exotic == "Native") %>% group_by(species) %>% summarise(total_abundance = sum(sp.n))

```

get transects meta data

```{r}
transect_mdata<-medata_19 %>% select(1:22) %>% distinct()
```


(3) keep trait for natives only (>19)


```{r}
species_traits_native<-species_traits %>% filter(exotic=="Native")

species_traits_native$exotic<- NULL
```


convert to gower distance matrix and save in "data.matrix" format


```{r include=FALSE}

# add the species as row names

my_traits <- species_traits_native %>% filter(species %in% natives_pool$species)%>% column_to_rownames(var="species") 


traits_gower_dist <- gowdis(my_traits)
traits_gower_matrix<-data.matrix(traits_gower_dist)

keep <- rownames(traits_gower_matrix)

```

(4) create the 100 null communities using the `create_null_community()` function

```{r message=FALSE, warning=FALSE}

set.seed(1)

null_medata_list <- replicate(100,create_null_community(), simplify = FALSE)

identical(null_medata_list[[1]], null_medata_list[[2]]) # should be false


```

# Run TD and FD calcultions

TD calc

```{r}

# results_td_list <- lapply(null_medata_list,function(x) estimate3D(
#       x,
#       diversity = "TD",
#       q = c(0, 1, 2),
#       datatype = "abundance",
#       base = "coverage",
#       level = c_min))

#save(results_td_list, file = here("3_diversity tables", "results_td_list.Rdata"))

load(here("3_diversity tables", "results_td_list.Rdata"))

# connect to the medata


meta.data<-transect_mdata %>% 
  mutate(Assemblage = paste(Site,enforcement,tmean,sep="_"),.after = Site) %>%  
  drop_na(tmean) %>%
  select(Assemblage,Basin,Site,enforcement,protection,tmean) %>%
  distinct(Assemblage,.keep_all = T)



null_diversity_tables <- lapply(results_td_list, process_td_results)

#save(null_diversity_tables, file = here("3_diversity tables", "results_td_null_list.Rdata"))

```

calculate functional diversity for each null community

```{r}
# results_fd_test <- pblapply(null_medata_list, function(x) {
# 
#   site_id_fd <- make_fd_matrix(x)
# 
#   estimate3D(site_id_fd,
#              diversity = "FD",
#              q = c(0, 1, 2),
#              FDdistM = traits_gower_dist,
#              datatype = "abundance",
#              base = "coverage",
#              level = c_min)})

#save(results_fd_test, file = here("3_diversity tables", "results_fd_null_list.Rdata"))

load(here("3_diversity tables", "results_fd_null_list.Rdata"))
```

Process the FD nulls

```{r message=FALSE, warning=FALSE}
null_diversity_tables_FD <- lapply(results_fd_test, process_fd_results)
```


# Unite the FD and TD and orgenize

```{r}

null_abundance_diversity <- Map(left_join,null_diversity_tables,null_diversity_tables_FD)


null_abundance_diversity <- lapply(
  seq_along(null_abundance_diversity),
  function(i) {
    null_abundance_diversity[[i]] %>%
      mutate(iteration = i, .before = 1)
  }
)

null_abundance_diversity_df <- bind_rows(null_abundance_diversity)

null_abundance_diversity_df$site_id<-NULL

null_abundance_diversity_df<-null_abundance_diversity_df %>% rename( Native_biomass=mean_native_biomass,
                                             Invasive_biomass=mean_invasive_biomass)

null_abundance_diversity_df<-null_abundance_diversity_df %>% 
  relocate(Native_biomass,.after = FD_simpson)%>%
  relocate(Invasive_biomass,.after = FD_simpson) 


null_abundance_diversity_df<-null_abundance_diversity_df %>% 
  relocate(mean_total_biomass,.after = Invasive_biomass )%>% rename(Total_biomass = mean_total_biomass)


null_abundance_diversity_df$protection<-as.factor(null_abundance_diversity_df$protection)

#save(null_abundance_diversity_df, file = here("3_diversity tables", "null_abundance_diversity_df.Rdata"))
```

# Species based null model

set 'replace' instead of any name of invasive

```{r}
medata_replacement<- medata_19

medata_replacement<-medata_replacement %>% 
  mutate(species = case_when(exotic == "Invasive" ~ "replace",
                             exotic == "Native" ~ species))
```

for each unique_trans_id in medata_replacement, if there is species named replace, choose randomly a species from the native pool to replace it. however, it need to be species that do not already appear in the transect and when adding new species it need to be without replacement so the same species can not be choose more than once.

```{r message=FALSE, warning=FALSE}

replace_species_in_transect <- function(df) {
  
  n_replace <- sum(df$species == "replace")
  
  if (n_replace == 0) return(df)
  
  species_already_present <- df %>%
    filter(species != "replace") %>%
    pull(species) %>%
    unique()
  
  candidate_pool <- natives_pool %>%
    filter(!species %in% species_already_present)
  
  if (n_replace > nrow(candidate_pool)) {
    stop("Not enough native species available to replace invasives in one transect")
  }
  
  new_species <- sample(candidate_pool$species,
                        size = n_replace,
                        replace = FALSE)
  
  df$species[df$species == "replace"] <- new_species
  
  return(df)
}

create_species_null <- function() {

  medata_replacement_native <- medata_replacement %>%
    group_by(unique_trans_id) %>%
    group_modify(~replace_species_in_transect(.x)) %>%
    ungroup()

  return(medata_replacement_native)
}

set.seed(1)

null_species_list <- replicate(
  100,
  create_species_null(),
  simplify = FALSE
)

make_fd_matrix_species_null <- function(null_data) {
  
  site_id <- null_data %>%
    group_by(across(1:23)) %>%
    summarise(amount = sum(sp.n), .groups = "drop") %>%
    spread(species, amount, fill = 0) %>%
    mutate(site_id = paste(Site, enforcement, tmean, sep = "_"), .after = Site) %>%
    drop_na(tmean)
  
  site_id <- site_id %>%
    select(site_id, 23:ncol(site_id)) %>%
    group_by(site_id) %>%
    summarise(across(.fns = sum), .groups = "drop") %>%
    column_to_rownames(var = "site_id")
  
  site_id <- t(site_id)
  
  site_id_fd <- as.data.frame(site_id)
  
  site_id_fd <- site_id_fd %>%
    filter(row.names(site_id_fd) %in% keep)
  
  site_id_fd <- data.matrix(site_id_fd)
  
  return(site_id_fd)
}

null_species_fd_matrix_list <- lapply(null_species_list, make_fd_matrix_species_null)

# note - the sampling of new species was in the transect level while here i have pooling in the site_id level and there for the number of natives can change a but (i.e, chromis apperd in most transect in the site but was absent in on and thus could be added as replacement for invasive in this transect resulting in a different number of chromis along iterations)

```



```{r}
  # results_species_null_fd <- pblapply(null_species_fd_matrix_list, function(site_id_fd) {
  #   # estimate3D(site_id_fd,
  #            diversity = "FD",
  #            q = c(0, 1, 2),
  #            FDdistM = traits_gower_dist,
  #            datatype = "abundance",
  #            base = "coverage",
  #            level = c_min)})

#save(results_species_null_fd, file = here("3_diversity tables", "results_fd_null_species_list.Rdata"))

load(here("3_diversity tables", "results_fd_null_species_list.Rdata"))

```

```{r message=FALSE, warning=FALSE}
null_diversity_tables_FD_species <- lapply(results_species_null_fd, process_fd_results)

# for simplicity and script convenience i unite it here with the TD data from last analyses - will be later omitted!


null_species_diversity <- Map(left_join,null_diversity_tables,null_diversity_tables_FD_species)

# set iteration number

null_species_diversity <- lapply(
  seq_along(null_species_diversity),
  function(i) {
    null_species_diversity[[i]] %>%
      mutate(iteration = i, .before = 1)
  }
)

# convert to data frame

null_species_diversity_df <- bind_rows(null_species_diversity)

# remove redundant column

null_species_diversity_df$site_id<-NULL

# format data table

null_species_diversity_df<-null_species_diversity_df %>% rename( Native_biomass=mean_native_biomass,
                                             Invasive_biomass=mean_invasive_biomass)

null_species_diversity_df<-null_species_diversity_df %>% 
  relocate(Native_biomass,.after = FD_simpson)%>%
  relocate(Invasive_biomass,.after = FD_simpson) 


null_species_diversity_df<-null_species_diversity_df %>% 
  relocate(mean_total_biomass,.after = Invasive_biomass )%>% rename(Total_biomass = mean_total_biomass)


null_species_diversity_df$protection<-as.factor(null_species_diversity_df$protection)

# set TD indexes to NA

null_species_diversity_df$TD_richness<-rep(NA)
null_species_diversity_df$TD_shannon<-rep(NA)
null_species_diversity_df$TD_simpson<-rep(NA)

# save

#save(null_species_diversity_df, file = here("3_diversity tables", "null_species_diversity_df.Rdata"))
```

