---
title: "Clean 3 - i next for all species"
author: "TG"
date: "2026-05-14"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


# In this script:

i-next calculation to rarafy TD and FD diversity for native communities

*Input data:*


* `medata_multistressor.rdata` - medata clean for this project

* `species_traits_multistressor.rdata` - traits data clean for this project with imputations

* `site_biomass.rdata` - table of the native and total biomass per site

* input files are at `2_data_for_analyses` folder


*output data:*

* `diversity_table_all_species.rdata` - rarefied TD and FD diversities of native communities

* `results_FD_all_species.Rdata` - since it computational heavy this sub-table is also saved for easier down stream code fixes 

* outputs saved in `3_diversity tables` folder

## Load packeges

```{r}
library(iNEXT.3D)
library(tidyverse)
library(vegan)
library(plotrix)
library(sjPlot)
library(sjmisc)
library(gridExtra)
library(ggeffects)
library(vegan)
library(plotly)
library(FD)
library(mFD)
library(funrar)
library(here)

select<-dplyr::select

```


```{r}

load(here("2_data_for_analyses", "medata_multistressor.rdata"))
load(here("2_data_for_analyses", "species_traits_multistressor.rdata"))
load(here("2_data_for_analyses", "site_biomass.rdata"))

```



### (1) preper data - all species

convert medata to wide format 


```{r}


medata_all<-medata %>% group_by(across(1:23)) %>% summarise(amount = sum(sp.n)) %>% ungroup() # sum the individuals from the same species in each transect

medata_all_wide<-spread(medata_all,species,amount,fill = 0)


```


keep trait for all species, remove exotic column

```{r}

species_traits$exotic<- NULL

```


convert to gower distance matrix and save in "data.matrix" format


```{r include=FALSE}

# add the species as row names

my_traits <- species_traits %>%  column_to_rownames(var="species") 


traits_gower_dist <- gowdis(my_traits)
traits_gower_matrix<-data.matrix(traits_gower_dist)

```

# Diversity indeces

## Calculate the covrage at the site id level

site_id = combination of site,enforcement and temperature


```{r}
site_id<-medata_all_wide %>%
  mutate(site_id = paste(Site,enforcement,tmean,sep="_"),.after = Site) %>% drop_na(tmean)

# keep just the id_column and the species and unite the counts 

site_id<- site_id %>%
  select(site_id,23:ncol(site_id)) %>%
  group_by(site_id) %>%
  summarise(across(.fns = sum), .groups = "drop")

# move the id_column to be the rowname

site_id<-site_id %>% column_to_rownames(var = 'site_id')

# transpose so the species are in the rows and site in column

site_id<- t(site_id)

```

see which species are in the data but dont have traits

```{r}
test<-rownames(site_id)

test[which(!test %in% species_traits$species)]
```



Calculate the sampling coverage per site 


```{r}
covrage_info <- DataInfo3D(site_id,
                                 diversity = "TD",
                                 datatype = "abundance")

```


now Ill calculate the Cmin and Cmax

```{r}
c_min<-min(covrage_info$`SC(n)`)

c_min

c_max<-min(covrage_info$`SC(2n)`)

# the c_min of the only native community

c_min<-0.9540893
```

now ill calculate the TD and FD according to those coverage values

```{r}

results_td<- estimate3D(site_id,
          diversity = 'TD',
          q = c(0,1,2),
          datatype = "abundance",
          base = "coverage",
          level = c(c_min,c_max))

# add whether we used Cmin or Cmax

results_td<-results_td %>%
  mutate("sc_type"= case_when(SC == min(SC)~ "Cmin",
                              SC ==max(SC)~ "Cmax"),.after = SC)

# connect to the medata


meta.data<-medata_all_wide %>% 
  mutate(Assemblage = paste(Site,enforcement,tmean,sep="_"),.after = Site) %>%  
  drop_na(tmean) %>%
  select(Assemblage,Basin,Site,enforcement,protection,tmean) %>%
  distinct(Assemblage,.keep_all = T)

# join with the TD (normalized to covrage)

td.diversity<-left_join(results_td,meta.data)

# convert enforcement to factor

td.diversity$enforcement<-as.factor(td.diversity$enforcement)

```

convert to wide data set

use cmin from the native only diversity

c_min<-0.9540893

```{r}


diversity_table<-td.diversity

diversity_table<-diversity_table %>% filter(sc_type=="Cmin")%>%  select(c(11:15,SC,Assemblage,Order.q,qTD))

diversity_table<-diversity_table %>% mutate(Order.q = case_when(
  Order.q == 0 ~ "TD_richness",
  Order.q == 1 ~ "TD_shannon",
  Order.q == 2 ~ "TD_simpson"
))

diversity_table<- spread(diversity_table,key =Order.q,value = qTD)

diversity_table<-left_join(diversity_table,site_biomass)

```
### Functional diversity per site 

```{r}

# list of the species i have trait data for them

keep <- rownames(traits_gower_matrix)

# filter the data accordingly

site_id_fd<-as.data.frame(site_id)

site_id_fd <-site_id_fd %>% filter(row.names(site_id_fd) %in% keep)

# convert back to data.matrix

site_id_fd<-data.matrix(site_id_fd)


```


run the coverage analyses 

*note* -  take long time, better do over the night if need to re-run or load to output



```{r}
# results_FD <- estimate3D(site_id_fd,# dim 1:75,1:128
#           diversity = 'FD',
#           q = c(0,1,2),
#           FDdistM = traits_gower_matrix,# dim 1:75,1:75
#           datatype = "abundance",
#           base = "coverage",
#           level = c(c_min))


#save(results_FD, file = here("3_diversity tables", "results_FD_all_species.Rdata"))

load(here("3_diversity tables", "results_FD_all_species.Rdata"))

```



```{r}
results_FD<-results_FD %>%
  mutate("sc_type"= case_when(SC == min(SC)~ "Cmin",
                              SC ==max(SC)~ "Cmax"),.after = SC)

fd.diversity<-left_join(results_FD,meta.data)

# convert enforcement to factor

fd.diversity$enforcement<-as.factor(fd.diversity$enforcement)


# convert to wide format

fd_table<-fd.diversity

fd_table<-fd_table %>%  select(c(11:15,SC,Assemblage,Order.q,qFD))

fd_table<-fd_table %>% mutate(Order.q = case_when(
  Order.q == 0 ~ "FD_richness",
  Order.q == 1 ~ "FD_shannon",
  Order.q == 2 ~ "FD_simpson"
))

fd_table<- spread(fd_table,key =Order.q,value = qFD)

# Join with the diversity table


diversity_table <-left_join(diversity_table,fd_table)

```
last cleaning 

```{r}
diversity_table$site_id<-NULL

diversity_table<-diversity_table %>% rename( Native_biomass=mean_native_biomass,
                                             Invasive_biomass=mean_invasive_biomass,
                                             Total_biomass=mean_total_biomass)

diversity_table<-diversity_table %>% 
  relocate(Native_biomass,.after = FD_simpson)%>%
  relocate(Invasive_biomass,.after = FD_simpson) 


diversity_table<-diversity_table %>% 
  relocate(Total_biomass,.after = Invasive_biomass )

diversity_table$protection<-as.factor(diversity_table$protection)
  
```


add data origin to diversity table

```{r}
data_origin<-medata %>% distinct(data.origin,Site)

data_origin$data.origin[data_origin$data.origin=="azz_asi"]<-"Azzurro"
data_origin$data.origin[data_origin$data.origin=="azz_malt"]<-"Azzurro"

diversity_table<-left_join(diversity_table,data_origin)

diversity_table$data.origin<-as.factor(diversity_table$data.origin)

diversity_table<-diversity_table %>% 
  relocate(data.origin,.before = Basin )

```

# save sites diversity tables

```{r}
diversity_table_all<-diversity_table

#save(diversity_table_all, file = here("3_diversity tables", "diversity_table_all.rdata"))

```



